One-dimensional Classical Gas in a Harmonic Trap with Repulsive Interaction

### 2017-12-14 draft_06.tex
working on introduction

\documentclass[aps,preprintnumbers,onecolumn,amsmath,amssymb,floatfix,pra]{revtex4-1}

@@

Will write Introduction after everything else is written.

@@

@@

Our model for the interacting classical gas involves particles with a simple finite-range repulsive
interaction.  Two particles repel each other whenever they are within a distance $\sigma$ from each
other.  The force of repulsion is a constant, $F_0$, within this distance and zero when the distance
is larger. 
%
\begin{equation}
\Big| F(|x|) \Big| ~=~ \begin{cases}
F_0  &\quad |x|<\sigma\\
0  &\quad |x|>\sigma
\end{cases}
\label{eq:preparation1}
\end{equation}
where $|x|$ is the distance between two particles.  This equation describes the magnitude of the
interparticle force; the direction is always repulsive.

@@

In the dilute gas limit, there are few or no interactions occuring at most times.  Thus the energy
$E$ is dominated by the trap energy, i.e., $E\approx m\omega_0^2\sum_{i}x_i^2= Nm\omega_0^2R^2$, or,
in our units,
\begin{equation}
  E\approx NR^2 \, .
  \label{eq:E_and_R}
\end{equation}
This is a good approximation at large $F_0$ because, for very large interactions, the particles
behave has hard rods which bounce off each other very rapidly, so that at most instants one can
expect no collisions to be taking place.  It is also expected to be a good approximation at very
small $F_0$ since we can then simply neglect the interaction energy.  Thus, we will use this as a common
estimate for $R$ in terms of the energy.

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure*}[h]
\includegraphics[width=0.9\textwidth]{ZhiyuPictures/stationary_and_rotating_frame_t=1-5_5T.eps}
%
\caption{Upper (lower) panels are snapshots of the cloud configuration in phase space in the
  stationary (rotating) frame, at instants within the first trap period.  Here $N=50$, $F_0=100$,
  and $E=50000$.  }
\label{fig:Breathingfrequency2_0}
\end{figure*}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure*}
\includegraphics[width=0.9\textwidth]{ZhiyuPictures/rotating_frame_t=0-80T.eps}  
\caption{Phase space snapshots of the cloud dynamics viewed in the rotating frame, at longer
  timescales.  The broadening and rotation observed here are interaction effects: in the
  non-interacting gas the particles are stationary in this frame.}
\label{fig:Breathingfrequency2_1}
\end{figure*}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[tbph]
\centering
\includegraphics[width=0.9\textwidth]{ZhiyuPictures/two_particles_a_01.pdf}
\caption{Left: Effective potential for two-particle system.  Center and right: beats at internal
  energy $E_i=2$ and $E_i=5$.  The curves show the individual energies of the two particles.  Here
  $F_0=1$. In this case, the larger the $E_i$, the lower the beat frequency. }
\label{fig:thermalization2}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

With interactions, a term $F_0(1-|r|)\theta(1-|r|)$ is added, where $r$ is the relative
displacement; as shown in Figure \ref{fig:thermalization2}.  Now, the relative motion is no longer
harmonic.  When the internal energy is much larger than $F_0$, one could regard the resulting motion
has having a slightly different frequency, or a collection of frequencies whose center is shifted
slightly from $\omega_0$.  Since the center-of-mass motion is still of frequency $\omega_0$, we have
a superposition of slightly different frequencies, resulting in beating dynamics.  This is clearly
seen in the time evolution of individual energies shown in Figure \ref{fig:thermalization2}.  The
two panels correspond to different internal energy $E_i$ (defined as the total energy minus the
center-of-mass energy).  For larger $E_i$, the distortion of the effective potential (at constant
$F_0$) plays a smaller role in shifting the effective frequency of relative motion; hence the beat
frequency is smaller.

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{ZhiyuPictures/pair1_pre.eps}
\caption{Time evolution of single-particle energies in a system of five particles, showing a case of
  inefficient energy redistribution.  The energies of the two partiles with highest energy show a beating
  similar to the two-particle case.}
\label{fig:thermalization4}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

\subsection{Condition for relaxation}

The main obstruct to get an thermalized state is the low frequency beat mentioned before. As a
result, to achieve ergodic state, we have to avoid such low frequency oscillation. According to our
discussion before, the solution is to make internal energy of each two-particle pair not ``too
large". Though it is impossible to express the internal energy of every pair in terms of the total
energy $E$, we can estimate it by the average energy, which is $\frac{E}{N}$. At least they are of
the same order. The thermalization condition could be given by:
\begin{equation}
F_0\sigma\sim\frac{E}{N}
\label{eq:thermalizatiton condition}
\end{equation}

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[tbhp]
\centering
%\includegraphics[scale=0.2]{ZhiyuPictures/distribution1_scanF0_E=100_3.eps}
%\caption{Distribution scan $F_0$, $E=100$, $N=5$, $\sigma=1$ {\color{red}{Run again!}} }
% \includegraphics[width=0.32\textwidth]{ZhiyuPictures/N=5_energydistribution_500_400_Font16.eps} 
% \includegraphics[width=0.32\textwidth]{ZhiyuPictures/N=10_energydistribution_500_400_Font16.eps}
% \includegraphics[width=0.32\textwidth]{ZhiyuPictures/N=20_energydistribution_500_400_Font16.eps}
%
\includegraphics[width=0.85\textwidth]{ZhiyuPictures/Boltzmann_2panels_01-crop.pdf}
%
\caption{the energy distribution measured in $\sim 10000$ time units, as will be verified later, Green curves corresponds to Boltzmann distribution; Red curves are obviously non-Boltzmann, while the yellow ones are intermediate case, represent ``thermalization threshold". In three figures for different N, we choose E/N identical (=1000), and thermalization is reached at $F_0=500\sim1000$, which is consistent with our prediction in Eq.\ref{eq:thermalizatiton condition}}
\label{fig:thermalization5}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

%\begin{figure}
%\subfigure[N=10, $F_0=200$]{
%\begin{minipage}[b]{0.4\linewidth}
%\includegraphics[width=1\textwidth]{ZhiyuPictures/N=10_single_particle_energy_F0=200_Font12_s.eps} 
%\label{fig:thermalization6}
%\end{minipage}
%}

%\subfigure[N=10, $F_0=1000$]{
%\begin{minipage}[b]{0.4\linewidth}
%\includegraphics[width=1\textwidth]{ZhiyuPictures/N=10_single_particle_energy_F0=1000_Font12_s.eps} 
%\label{fig:thermalization7}
%\end{minipage}
%}

%\end{figure}

%% %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
%% \begin{figure}[h]
%% \subfigure[N=10, $F_0=200$]{
%% \begin{minipage}[b]{0.4\linewidth}
%% \includegraphics[width=1\textwidth]{ZhiyuPictures/N=10_single_particle_energy_F0=200_s_400_300_Font12.eps}
%% \label{fig:thermalization6}
%% \end{minipage}
%% }
%% \subfigure[N=10, $F_0=1000$]{
%% \begin{minipage}[b]{0.4\linewidth}
%% \includegraphics[width=1\textwidth]{ZhiyuPictures/N=10_single_particle_energy_F0=1000_s_400_300_Font12.eps} 
%% \label{fig:thermalization7}
%% \end{minipage}
%% }
%% \caption{Time evolution of the particles' energy $\epsilon$}
%% \end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

As is shown in Fig.\ref{fig:thermalization5} above, the critical point for reaching ``Boltzmann-like" distribution (later we will verify that it is Boltzmann distribution) is $F_0=500\sim1000$ (we always set $\sigma=1$), while the average energy $\sim 1000$, as we expected, they are of the same order. As a supplementary proof of our argument, Fig.\ref{fig:thermalization6} and Fig.\ref{fig:thermalization7} show the great difference of single-particle energy between two examples of below and above the thermalization threshold ($F_0=200$ $\&$ $F_0=1000$ for $N=10, E=10000$): in the former one there is always some particle maintained at high excitation, while in the latter the system probably goes to ergodic. To date, we have verified that the threshold of thermalization, which is $\frac{NF_0\sigma}{E}\sim1 $ , and corresponding feature of their energy distribution in two side of the threshold.

One may think that this condition only means that the energy transfer in every collision is much smaller than the energy interval. But this is not the whole story. The essential difference between these two regime lies in whether the energy transfer in every collision is correlated: In thermalizable regime, the energy transfer in every collision is much smaller than the energy interval, which definitely slows down the energy transfer. What is more, because of this, the particles is less interferred by other particles so that two-particle analysis survives. It means that for a particle pair, the energy transferred in this collision is correlated with the one in the next time. In this manner, each particle in the pair will pick up some energy in several collisions, and then return it back to its partner -- which is just what the low frequency beat effect tells us. Thus the particles energy is localized in certain value. On the contrary, if we go to non-thermalizable regime, the energy transfer is so fast that for each particle don't remember their partners. As a result, the energy transferred in every collision between two particles are no longer correlated so that our two-particle analysis breaks down. Instead, the energy transfer is completely random, with a non-zero amplitude. On the single-particle energy diagram, what we expected to see is a ``one dimensional random walk" ,so energy levels quickly spread around the diagram.

To sum up, the thermalization threshold not only tells us whether the amplitude of energy transfer in every collision is small enough compared to particles' energy, but also shows whether the ``correlation time" of particle pair is much longer than the characteristic time of the system (the oscillation period).

\subsection{Verifying Boltzmann distribution}
Thermalization could have different definition. In our system, the ``thermalization" we want to find means ``losing all the memory of initial state". One of the most important information of initial state is energy distribution.

For an isothermal system, the energy distribution of the whole system follows the Boltzmann distribution in equilibrium state. However, there is not any obvious conclusion about the distribution for an isolated system where the total energy is conserved. But intuitively, one will expect that if we measure the energy of a subsystem, e.g. one particle, then we will get a Boltzmann distribution because the rest part of the system can be considered as a bath for this particle. The temperature of this isolated system is defined according to $Nk_BT=E$. It is evident that this argument only hold when the energy of the single particle is not too big --- if one particle take up 50\% of the total energy, the rest could no longer be thought of as a good bath.

In the first picture below, we verify this argument by measuring the energy distribution of every particle at N=10 for different F0 at E=1000 (parameters here satisfy thermalizing condition). In the other two, we tested different N.

%\begin{figure*}
%%\includegraphics[scale=0.5]{ZhiyuPictures/distribution_at_N=5_F0=10000_pre.eps}
%%\caption{Energy distribution, N=5, F0=10000}
%%\label{fig:thermalization8}
%\subfigure[N=10, $F_0=1\times10^4$]{
%\begin{minipage}[b]{0.1\textwidth}
%\includegraphics[scale=0.1]{ZhiyuPictures/boltzmann_N=10_F=10000_pre_rev.eps}
%\caption{Energy distribution, N=10, $F_0=1\times10^4$}
%\label{fig:thermalization9}
%\end{minipage}
%}

%\subfigure[N=20, $F_0=1\times10^4$]{
%\begin{minipage}[b]{0.1\textwidth}
%\includegraphics[scale=0.1]{ZhiyuPictures/boltzmann_N=20_F=10000_pre_rev.eps} 
%\caption{Energy distribution, N=20, $F_0=1\times10^4$}
%\label{fig:thermalization10}
%\end{minipage}
%}

%\end{figure*}

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[h]
\subfigure[N=10, $F_0=1\times10^4$]{
\begin{minipage}[b]{0.4\linewidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/boltzmann_N=10_F=10000_500_400_Font16.eps}
\label{fig:thermalization9}
%\caption{Energy distribution, N=10, $F_0=1\times10^4$}
\end{minipage}
}
\subfigure[N=20, $F_0=1\times10^4$]{
\begin{minipage}[b]{0.4\linewidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/boltzmann_N=20_F=10000_500_400_Font16.eps} 
\label{fig:thermalization10}
%\caption{Energy distribution, N=20, $F_0=1\times10^4$}
\end{minipage}
}
\caption{Fitting energy distribution $\rho(\epsilon)$ with Boltzmann distribution}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

>>>>

\documentclass[onecolumn,pra]{revtex4-1}

@@

The non-equilibrium dynamics of isolated systems has been the subject of a large volume of work in
recent years.  In particular, in the quantum context, the study of unitary quantum evolution of
many-body and few-body systems has undergone an explosive growth.  This interest was partly fueled
by the availability of experimental systems, e.g., trapped ultracold atomic and ionic systems, where
non-dissipative dynamics can be studied explicitly due to excellent isolation (CITE).  There are
also fundamental reasons for studying dynamics in isolation: recent activity has led to advances in
the understanding of basic principles connecting quantum dynamics, quantum chaos and the emergence
of statistical mechanics and thermalization (CITE).  For classical many-body systems, explorations
of the connections between microscopic dynamical rules, chaos, ergodicity and statistical mechanics
have a much longer history, dating back to the time of Boltzmann and Poincare.
\cite{Dorfman_book_1999, Dumas_book_KAMstory, Boltzmann_legacy_book, EckmannRuelle_RMP85}.

Motivated by questions and issues arising in the field of quantum many-body and few-body dynamics,
in this work we will study the classical dynamics of a collection of interacting particles confined
in a one-dimensional harmonic trap.  The analogs of two questions, studied intensively in the
quantum literature, will be investigated: (1) the influence of interactions on the breathing mode
frequency of the trapped gas, and (2) the relaxation and possible thermalization of the finite
system under its own evolution, in the absence of any external baths or dissipation mechanisms.

Experiments with cold atoms, which have had an essential motivational role for the current interest
in many-body quantum dynamics, are generally performed in a trapping potential, either optical or
magnetic.  As a result, in the last two decades many questions of many-body quantum physics have
been reformulated in the presence of a harmonic trap.  The presence of a harmonic trap often results
in new dynamical phenomena; for example the motion of vortices in a trapped condensate (CITE) has
new features (such as precession) compared to vortex motion in uniform condensates (CITE).  Trapping
also results in new collective phenomena which have no analog in uniform situations: for example,
collective modes like dipole, breathing and quadrupole modes are specific to trapped many-body
systems.  Dipole modes (center of mass oscillations) and breathing modes (size oscillations) appear
even for one-dimensional trapped systems; higher-dimensional systems display additional, more
complicated modes like quadrupole and scissors modes.  Because of the omnipresence of trapping
potentials in cold-atom experiments, dipole and breathing mode dynamics are pervasive across the
field: These modes have been intensively studied and used for diagnostic purposes since the early
days of research with trapped quantum-degenerate gases, and continue to generate interest today
\cite{Naegerl_Science2009, Esslinger_exp_PRL2003, collective_modes_early_papers,
  Tschischik_BHbreathing_PRA_2013, KroenkeSchmelcher_BM, Bonitz_Review, Bouchoule_PRL2014,
  our_superTG_PRA_2015, STG_Astrakharchik_PRL2005, STG_Chen-etal_PRA2010,
  br_mode_long_range_interactions, 1D_breathing_mode_recent, KohnMode, Cominotti_2015,
  ZarembaTowers_PRA2014, Wetterich_JPB2011, fermion2Dbreathing, Stringari_lowDcollective_PRA15,
  Mazets_1Dbreathing_EPJD11}.  (CITE).
%
In one of the best-known experiments on quantum many-body dynamics,
Ref.~\cite{NewtonsCradle_Nature06}, the dynamics of an integrable system (Lieb-Liniger bosons) was
seen to have ultraslow relaxation in a trap, supposedly due to integrability.  This experiment has
led to continuing efforts to understand how a trap breaks integrability (CITE).

The presence of harmonic trapping is thus a paradigmatic aspect in the field of quantum many-body
dynamics.  This motivates our study of classical dynamics of trapped interacting systems.  Similarly
motivated by quantum dynamics, classical trapped dynamics of hard rods has recently been studied in
Refs.\ \cite{DoyonSpohn_JSTAT17, Moore_arXiv17.10}, focusing on integrability-breaking effects of
the trap.  In this work, we consider trapped classical particles which interact via a simple
finite-range potential between any pair of particles: a constant repulsive force acts whenever the
particles are less than a certain distance apart.  The interaction strength can be tuned, leading to
different regimes of behavior.

Our first theme will be a study of the breathing mode of the classical interacting gas, and the
effect of interactions on the frequency of the breathing mode.  In the quantum case, this has long
been a topic of interest for both fermions and bosons.  For bosons with short-range interactions,
the situation has been studied in one, two and three dimensions.  At zero interactions, the
breathing mode is twice the trapping frequency $\omega_0$ (assuming isotropic trapping) in each
case.  A mean field (Gross-Pitaevskii) calculation gives the frequency to be $\sqrt{D+2}$ times the
trapping frequency, where $D$ is the spatial dimension.  For three-dimensional (3D) traps, the
mean-field description can be expected to be qualitatively correct, thus the breathing mode
frequency is expected to change monotonically from $2\omega_0$ at zero interactions to
$\sqrt{5}\omega_0$ at large interactions.  For 2D traps, the breathing frequency is exactly
$2\omega_0$ for any interaction strength \cite{PitaevskiiRosch_PRA97}.  This remarkable result
follows from a symmetry in 2D.  Classical trapped particles with scale-invariant interaction
potentials also have breathing frequency $2\omega_0$ independent of interaction strength
\cite{PitaevskiiRosch_PRA97}.  The interaction dependence in 1D (trapped Lieb-Liniger gas) is also
remarkable: as the interaction is increased, the breathing frequency first decreases from
$2\omega_0$ to the mean-field prediction $\sqrt{3}\omega_0$, and then at larger interactions
increases again, returning to $2\omega_0$ in the infinite interaction (Tonks-Girardeau) limit
\cite{Naegerl_Science2009, Tschischik_BHbreathing_PRA_2013, 1D_breathing_mode_recent,
  Bouchoule_PRL2014, KroenkeSchmelcher_BM}.
%
The breathing mode has also been investigated extensively for fermionic systems with contact
interactions \cite{GrimmSmith_unitaryfermions_PRA08, fermion2Dbreathing,
  Stringari_lowDcollective_PRA15}, and also for fermionic and bosonic systems with long-range
interactions \cite{br_mode_long_range_interactions, Bonitz_Review}.  The breathing mode has also
been studied for a high-temperature trapped gas where a Boltzmann-equation description is valid,
both theoretically \cite{Guery-Odelin1999} and experimentally
\cite{Cornell_classsical_3Dexpt_NatPhys15}.

In this work, we use an explicit microscopic model, rather than a Boltzmann equation approach.
Unlike the quantum case, the harmonic potential does not impose a length scale or an energy scale.
The interaction range can thus be scaled away by a redefinition of length, i.e., by measuring
lengths in units of the interaction range.  There are only two important parameters determining the
system behavior, namely the interaction strength and the energy per particle.  We find that the
breathing mode frequency has a non-monotonic dependence on the interaction parameter.  At small
interactions, when the particles pass through each other, the effect of collisions is to slow down
the dynamics, so that the breathing frequency drops below the non-interacting value $2\omega_0$.  At
very large interactions, the particles bounce off each other, a process which speeds up the size
oscillations, resulting in the breathing frequency being larger than $2\omega_0$.

Our second theme is the process of relaxation at long timescales, and possible thermalization.  This
topic is also motivated by intense recent research in the quantum context.  In quantum isolated
systems, thermalization is now understood in terms of the so-called eigenstate thermalization
hypothesis (ETH), which postulates that in ergodic systems the expectation values of observables in
individual eigenstates depend ony on the eigenenergy, and hence represents the thermal value (CITE).
The connection to statistical mechanics is particularly difficult for explicitly finite systems, in
which case ETH is to be understood in terms of finite size scaling (CITE).  In classical systems,
the connection between microscopic dynamics and statistical mechanics has been continually studied
since the 19th century.  It is generally understood that systems with positive Lyapunov exponents

@@

The words ergodic, relaxation and thermalization

Loss of memory of initial state --- our two definitions

@@

Our model for the interacting classical gas involves particles with a simple finite-range repulsive
interaction.  Two particles repel each other whenever they are within a distance $\sigma$ from each
other.  The force of repulsion is a constant, $F_0$, within this distance and zero when the distance
is larger. 
%
\begin{equation}
\Big| F(x) \Big| ~=~ \begin{cases}
F_0  &\quad |x|<\sigma\\
0  &\quad |x|>\sigma
\end{cases}
\label{eq:preparation1}
\end{equation}
where $|x|$ is the distance between two particles.  This equation describes the magnitude of the
interparticle force; the direction is always repulsive.

@@

In the dilute gas limit, there are few or no interactions occuring at most times.  Thus the energy
$E$ is dominated by the trap energy, i.e., $E\approx m\omega_0^2\sum_{i}x_i^2= Nm\omega_0^2R^2$, or,
in our units,
\begin{equation}
  E\approx NR^2 \, .
  \label{eq:E_and_R}
\end{equation}
%
This approximation is good when the average distance betwen particles is much larger than the
interaction range $\sigma=1$, i.e., whenever $R\gg1$, which is the regime we consider dynamics in.
Another way of describing this dilute-gas regime is that the gas is far from the ground state
(described in the next subsection).  Neglecting the interaction energy to estimate $E$ is reasonable
even at very large interactions ($F_0\gg E/N$), because for very large interactions, the particles
behave as hard rods which bounce off each other very rapidly, so that at most instants one can
expect no collisions to be taking place.  It is also expected to be a good approximation at very
small $F_0$ since we can then simply neglect the interaction energy.  Thus, we will use this as a
common estimate for $R$ in terms of the energy.

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure*}[tbph]
\includegraphics[width=0.9\textwidth]{ZhiyuPictures/stationary_and_rotating_frame_t=1-5_5T.pdf}
%
\caption{Upper (lower) panels are snapshots of the cloud configuration in phase space in the
  stationary (rotating) frame, at instants within the first trap period.  Here $N=50$, $F_0=100$,
  and $E=50000$.  }
\label{fig:Breathingfrequency2_0}
\end{figure*}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure*}
\includegraphics[width=0.9\textwidth]{ZhiyuPictures/rotating_frame_t=0-80T.pdf}  
\caption{Phase space snapshots of the cloud dynamics viewed in the rotating frame, at longer
  timescales.  The broadening and rotation observed here are interaction effects: in the
  non-interacting gas the particles are stationary in this frame.}
\label{fig:Breathingfrequency2_1}
\end{figure*}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[tbph]
\centering
\includegraphics[width=0.9\textwidth]{ZhiyuPictures/two_particles_a_01.pdf}
\caption{Two-particle dynamics.  Left: Effective potential for relative motion.  Center and right:
  evolution of the individual energies of the two particles.  The two panels correspond to internal
  energy $E_i=2$ and $E_i=5$; the interaction is $F_0=1$.  The beat frequency is lower for larger
  $E_i$. }
\label{fig:thermalization2}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

With interactions, a term $F_0(1-|r|)\theta(1-|r|)$ is added, where $r$ is the relative
displacement; as shown in Figure \ref{fig:thermalization2}.  Now, the relative motion is no longer
harmonic.  When the internal energy is much larger than $F_0$, one could regard the resulting motion
as having a slightly different frequency, or a collection of frequencies whose center is shifted
slightly from $\omega_0$.  Since the center-of-mass motion is still of frequency $\omega_0$, we have
a superposition of slightly different frequencies, resulting in beating dynamics.  This is clearly
seen in the time evolution of individual energies shown in Figure \ref{fig:thermalization2}.  The
two panels correspond to different internal energy $E_i$ (defined as the total energy minus the
center-of-mass energy).  For larger $E_i$, the distortion of the effective potential (at constant
$F_0$) plays a smaller role in shifting the effective frequency of relative motion; hence the beat
frequency is smaller.

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{ZhiyuPictures/pair1_pre.eps}
\caption{Time evolution of single-particle energies in a system of five particles, showing a case of
  inefficient energy redistribution.  The energies of the two particles with highest energy (4 \& 5)
  show a clear beating similar to the two-particle case, as does the pair 2 \& 3, although less
  visible.  The `unpaired' particle 1 performs slow beating dynamics with the center of mass of the
  2 \& 3 pair.}
\label{fig:thermalization4}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

\subsection{Relaxation condition and the Boltzmann distribution}

Our consideration of few-particle motion has shown that relaxation is slow when the internal energy
of pairs is large compared to $F_0$.  This allows us to conjecture a condition for relaxation in
reasonable time.  Although it is impossible to express the internal energy of every pair in terms of
the total energy $E$, we can estimate the typical internal energy by the average energy,
$\frac{E}{N}$. At least they are of the same order.  Thus we have the condition
\begin{equation}
F_0 \gtrsim \frac{E}{N}
\label{eq:thermalizatiton_condition}
\end{equation}
for relaxation.  This is the same condition we obtained for the breathing frequency to be increased
(rather than decreased) by the interaction.

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[tbhp]
\centering
%\includegraphics[scale=0.2]{ZhiyuPictures/distribution1_scanF0_E=100_3.eps}
%\caption{Distribution scan $F_0$, $E=100$, $N=5$, $\sigma=1$ {\color{red}{Run again!}} }
% \includegraphics[width=0.32\textwidth]{ZhiyuPictures/N=5_energydistribution_500_400_Font16.eps} 
% \includegraphics[width=0.32\textwidth]{ZhiyuPictures/N=10_energydistribution_500_400_Font16.eps}
% \includegraphics[width=0.32\textwidth]{ZhiyuPictures/N=20_energydistribution_500_400_Font16.eps}
%
\includegraphics[width=0.85\textwidth]{ZhiyuPictures/Boltzmann_2panels_01-crop.pdf}
%
\caption{The distribution of single-particle energies measured in $\sim 10000$ time units. Green
  curves obey the relaxation condition \eqref{eq:thermalizatiton_condition}, and yellow/orange
  corresponds to smaller $F_0$.  In each panel (different $N$), we choose $E/N$ identical ($=1000$),
  so that the threshold interaction between the two behaviors is  $F_0\sim1000$. }
\label{fig:thermalization5}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

Because the model is expected to be ergodic for $N>2$ particles, we expect the same condition to
lead to thermalization.  In fact, the proposed condition works even quantitatively, as we show in
Figure \ref{fig:thermalization5}.  Here, the distribution of energies among the different particles
is presented.  Given our small particle number, a single snapshot of the energies will not yield
enough statistics to investigate the distribution of single-particle energies.  Therefore,
single-particle energies are recorded at once every time unit (i.e., at time intervals of
$1/omega_0$) as the simulation evolves, up to around $t\sim10^4$.  The distribution of these
observed values then shows whether or not the system has thermalized to a Boltzmann distribution.
In the two panels of Figure \ref{fig:thermalization5}, we show the distributions obtained for a
$N=5$ system and a $N=20$ system.  In both cases, we choose the total energy to be $E=1000N$.  In
accordance with our conjectured condition, we find that the observed single-particle distributions
are qualitatively different for $F_0\gtrsim10^3$ and for smaller interactions.  When the
``thermalization condition'' $F_0\gtrsim E/N$ is satisfied, the distribution is roughly exponential
($\sim e^{-\beta\epsilon}$.  For smaller $F_0$, thermalization in this sense is not seen in the
single-particle energy distribution.  Since the systems are expected to be ergodic, it is assumed
that the smaller-$F_0$ systems will also eventually show a Boltzmann distribution of energy, but
that this appears only at (much) longer timescales.

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[tbph]
\includegraphics[width=0.8\textwidth]{ZhiyuPictures/Boltzmann_log_2panels_01-crop.pdf}
%
\caption{\label{fig:thermalization9} Complaring the single-particle energy distribution
  $\rho(\epsilon)$ with the Boltzmann distribution.  The energy values all satisfy the relaxation
  condition \eqref{eq:thermalizatiton_condition}.  The Boltzmann distribution $\propto
  e^{-\beta\epsilon}$, with the temperature defined as $\beta^{-1}= E/N$, is the thick dashed line.}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

In several `thermalizing' situations, $F_0\gtrsim E/N$, we display the agreement with the Boltzmann
distribution in Figure \ref{fig:thermalization9}.  In each panel, the calculated distributions from
simulations is plotted against the Boltzmann distribution.  The overall agreement shows that the
energy is well-thermalized among the particles on average within the timescales under consideration.

There are some small deviations from the Boltzmann distribution visible in Figure
\ref{fig:thermalization9}.  Most of the deviation is probably due to numerical noise, and
unsurprising given our small system sizes.  However, some deviation is also expected due to the
contribution of the density of states (DOS).  The probability of finding an energy is actually
proportional to the Boltzmann exponential factor multiplied by the DOS.  For the simple harmonic
oscillator, the DOS is constant.  The interaction deforms the energy shell in the small region
$|x_i-x_j|<\sigma$; so that the DOS is no longer a constant.  The effect of interaction is
conspicuous in low energy part: when interaction is weak, particles prefer to stay near zero energy
according to Boltzmann distribution, but when interaction grows, they are no longer allowed to
accumulate near zero energy, in another word DOS at low energy decreases with the increase of
interaction. Therefore, counts at the low energy part of Fig.\ref{fig:thermalization9} is slightly
lower than the Boltzmann prediction.  The correction to the density of states is not simple to
calculate explicitly.  Since we are in the regime of a dilute gas ($\sigma\ll R$), the effect is
very small.

\subsection{Lyapunov Exponents}

The general idea that thermalization is faster for larger $F_0$ can be seen through the Lyapunov
exponents.  For high-dimensional systems, there is a spectrum of exponents which manifests the
instability of the trajectory along each direction. The largest Lyapunov exponent (LLE) reflects
the shortest time scale for the system lose memory of the initial state.  We follow the method
described in Ref.~\cite{EckmannRuelle_RMP85}.  If the calculation is carried out upto sufficiently
late times, the LLE for an ergodic system is uniquely defined and independent of the starting
state.  We would like to deal with parameter regimes both within and outside our relaxation
condition $E<F_0N$ and $E>F_0N$, and in the latter case we have seen that ergodicity does not become
apparent at reasonable timescales.  Therefore, we carry out the computation of the LLE upto
$t\sim10^3$, which means that we can only get initial-configuration-dependent estimates.  The
collection of these estimates forms a distribution which we show in Figure
\ref{fig:LLEdistribution1}.

%% The Lyapunov Exponents is a good tool to quantify the time scale of ``thermalization". Lyapunov
%% Exponents(LE) describes how fast one orbit diverge from its nearby orbits in $\Gamma$-space. But for
%% high dimensional system, LE is a spectrum that manifest the instability of the trajectory along each
%% direction. The largest Lyapunov exponents (LLE) reflects the shortest time scale that system lose
%% its memory. Each point along one trajectory in $\Gamma$-space could have different LLE value, but
%% the distribution of LLE is dependent on macroscopic parameters. As is shown in
%% Fig.\ref{fig:LLEdistribution1}, we measured the LLEs along our trajectory and plot its distribution
%% as a function of total energy.

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[tbph]
\centering
%\includegraphics[scale=0.4]{ZhiyuPictures/LLEdistribution_1_11_pre_hot1_screenshot.png}
\includegraphics[scale=0.4]{ZhiyuPictures/LLEdistribution_1_11_pre_hot2_screenshot.png}
\caption{The distribution of largest Lyapunov exponents, for $N=5$, $F_0=1000$, at different
  energies.  The color indicates the probability of a measured LLE value being found in a bin; the
  bins are of size 0.01}\label{fig:LLEdistribution1}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

%% \begin{figure}[h]
%% \centering
%% \includegraphics[scale=0.4]{ZhiyuPictures/LLEdistribution_1_11_pre_bone1_screenshot.png}
%% \includegraphics[scale=0.4]{ZhiyuPictures/LLEdistribution_1_11_pre_bone2_screenshot.png}
%% \caption{``bone". top(bottom):modified(reversed)}
%% \end{figure}

The LLE decreases with the growth of energy. The critical part is near $E=5000$. The most probable
value of LLE goes down from around 1 to around 0.1,which indicates that the shortest time scale
becomes longer than an oscillation period when $E\gtrsim5000$.  This is consistent with our
predicted thermalization threshold.  As $E$ increases far beyond the thermalization threshold value,
the LLE distribution is close to zero, suggesting a divergence of relaxation time.
%
We have thus shown a correspondence between energy thermalization and LLE.

\subsection{Shape of distribution in single-particle phase space}\label{section:Shape}

We have studied thermalization in terms of energy.  More generally, thermalization may be taken to
mean that a many-body system loses memory of its initial state.  In the single-particle phase-space
picture, Figures \ref{fig:Breathingfrequency2_0} and \ref{fig:Breathingfrequency2_1}, our initial
state is very special: The particles are lined up along the $X$ axis.  The effect of interaction is
to distort this line toward a circularly symmetric distribution.  Thus, we would consider that the
memory of the initial state is lost when the distribution of points in the $X$-$P$ plane are not
elongated in any one direction.
